Quantifying Infants' Everyday Restrained Experiences in the Home Using Wearable Inertial Sensors

### 2025-11-21 manuscript_restrained_classification validation.qmd
fixed typo

abstract: "Physical restraint—including being held, carried, and restrained in devices—is a common feature of infants’ everyday lives. However, previous survey-based and video-based methods cannot simultaneously provide continuous, full-day accounts of infant restraint experiences. This study developed and validated a machine learning model to quantify infants’ restraint time moment-to-moment across the full day in the home environment using wearable sensor movement data derived from an existing data set. The data set includes 142 home-visit sessions from 65 infants, with 29 younger infants aged 4-7 months, and 36 older infants aged 11-14 months. We annotated infants’ restrained states in the first 1.5-hour video recordings of each session as ground truth labels. The supervised machine-learning model achieved high accuracy (89%) and substantial kappa agreement (κ = .73) compared with human-coded ground truth. The model presented a slight bias towards overestimating unrestrained periods compared to restrained periods, but the bias was mitigated when using a longer data-aggregating window length. The model also showed convergent validity by corroborating prior studies that showed an age-related decrease in infants' overall restraint time throughout the day. In short, the current study demonstrated the feasibility of using wearable sensors to quantify infants’ real-world restraint experiences, offering a new tool for studying how daily restraint influences early development."

@@

bibliography: [references.bib]

@@

Physical restraint is a common feature in infants’ everyday lives. Infants are routinely held or carried by caregivers, and infants are often restrained in devices such as highchairs, walkers, and strollers [@Birken2015s]. As an externally regulated aspect of infants’ daily life, restraint shapes infants’ experiences based on caregivers’ decisions--caregivers, not infants, choose when infants are restrained. Restraint limits infants’ physical activities and might hinder infants’ motor development [e.g., @Pin2007z]. Restraint also alters experiences that are relevant for perceptual and cognitive development by determining what is accessible in the physical world. For example, restraint is associated with fewer object holding experiences [@Franchak2024l], reduced amount of language input [@Malachowski2023a], and increased opportunities for joint attention [@Karasik2025h]. Therefore, characterizing infants’ restrained experiences helps profile infants’ daily learning contexts and situate their motor, perceptual, and cognitive development within everyday experiences.

@@

Despite different approaches, prior works converge to show that restraint is common in infant daily life, and that the time infants spend restrained changes over development. In terms of caregiver restraint, 2- to 6-month-old infants spent around 40% of their awake time being held or carried [@Malachowski2023a; @Siddicky2020f]. Older 10- to 13-month-olds spent around 19% of their awake time being held or carried [@Franchak2024l]. In terms of device restraint, 4- to 6-month-olds spent around 20% of awake time in seating devices, and 35% in placement with physical support [@Callahan1997s; @Malachowski2023a]. Older 10- to 13-month-olds spent around 27% of awake time in restraining furniture or devices [@Franchak2024l]. Additionally, a study that compared across ages also confirmed that infants’ overall restrained time decreased from 2 to 6 months of age [@Carson2022n].

@@

Researchers have used survey methods, including traditional retrospective report and ecological momentary assessment (EMA), to quantify infants’ restrained time, but survey methods have limitations. Retrospective parent report asks parents to estimate their infant’s average daily restrained time over the past few weeks or month [@Abbott2001u; @Bartlett2003w; @Callahan1997s; @Carson2022n; @Hesketh2015i; @Karasik2018t; @Siddicky2020f]. Despite the widespread use of retrospective reports, they lack precision by relying on parents’ memory—it is difficult to recall and sum every bout of restraint practices and then report an overall time estimation [@Bradburn1987g]. There are exceptions where caregivers were asked to complete structured daily diaries reflecting on the previous day’s restraint practices in short intervals (e.g., every five minutes) [@Karasik2018t; @Majnemer2005j]. However, this method might still be sensitive to memory loss and thus not as accurate in providing data. EMA survey techniques reduce caregivers’ memory burden by making momentary phone-survey prompts to caregivers multiple times (e.g., 5, 10, or 12 times) randomly throughout a day [@Franchak2019o; @Franchak2024l; @Malachowski2023a]. Despite the benefits of gathering immediate observations rather than retrospective reports, EMA cannot provide continuous time-series data. Thus, a method is needed that provides time-sequence information without relying on parent report.

Video recording and human annotation overcome the above drawbacks of survey-based methods, but video is subject to other limitations. One disadvantage of video observation is being obtrusive. The presence of a researcher in the room may affect the caregiver’s and infant’s behaviors. Another limitation is that video recording and video annotation are both labor-intensive and time-consuming, which limits its use in large-scale or long-duration studies. Indeed, prior work that videotaped and manually annotated infants’ restraint had short durations of recording (e.g., half an hour in @Springfield2025p, and one hour in @Wang2025v). Within the short duration of recording, researchers tend to record the period when infants are awake and active. In contrast, periods when the infants are restrained, like seated in front of the TV or seated in highchairs, would likely be skipped. Thus, video recording is subject to sampling bias and might not be representative of infants’ whole-day experiences.

@@

However, one of the challenges researchers continue to face is the reduced model performance in naturalistic home settings [@Airaksinen2024z; @Franchak2024e; @Yao2019q], compared to more controlled laboratory environments [@Airaksinen2020j; @Franchak2021k]. Indeed, infants may exhibit more complex and unexpected movements at home versus in lab, which requires the models to be able to generalize to scenarios that go beyond the models’ training data set. Moreover, since no prior work has used wearable sensors to detect infants’ overall restrained periods, the current study addresses a unique detection problem.

@@

The third step was to train and validate supervised machine learning models that classify the motion features into restrained status. We applied the random forest algorithm (Breiman, 2001) in modeling using the *DecisionTree* package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as a testing set and trained the model using the other sessions, to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked[,]{.underline} and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

@@

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). This reproducible manuscript, including the model evaluation, figures, and statistical analyses, was made with Quarto version 1.5.57 (Allaire et al., 2024) using R version 4.5.0 (RStudioTeam, 2024) with the R packages tidyverse [@tidyverse], flextable [@flextable], and papaja [@papaja], as well as the apa quarto template (Schneider, n.d.). Human annotation results, which were used for model training and validation, together with processing scripts are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

@@

```{r fig-cibtwctg}
#| fig-cap: Differences in sensitivity and PPV between restrained and unrestrained prediction according to window length
#| fig-width: 8
#| fig-height: 5
#| apa-note: Each vertical line indicates the 95%CI of the metric's difference between restrained prediction and unrestrained prediction. The dot on each line denotes the mean of the difference (restrained - unrestrained).

@@

@tbl-indexbetweenmodels shows that all window lengths reached high accuracy (`r apa_num(table_btw_mdl[1,3])` for 4 s, `r apa_num(table_btw_mdl[1,4])` for 16 s, and `r apa_num(table_btw_mdl[1,5])` for 30 s), and good kappa values (`r apa_num(table_btw_mdl[2,3])` for 4 s, `r apa_num(table_btw_mdl[2,4])` for 16 s, and `r apa_num(table_btw_mdl[2,5])` for 30 s). All window lengths presented high sensitivity—how often the model correctly detected when the infant was actually restrained or unrestrained—ranging from `r apa_num(min(table_btw_mdl[3,3:5]))` to `r apa_num(max(table_btw_mdl[3,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[4,3:5]))` to `r apa_num(max(table_btw_mdl[4,3:5]))` for unrestrained prediction. All window lengths had high positive predictive value (PPV)—how often the model’s predictions of being restrained or unrestrained were correct—ranging from `r apa_num(min(table_btw_mdl[5,3:5]))` to `r apa_num(max(table_btw_mdl[5,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[6,3:5]))` to `r apa_num(max(table_btw_mdl[6,3:5]))` for unrestrained prediction. All window lengths also had high F1 scores—a balanced measure that combines sensitivity and PPV—ranging from `r apa_num(min(table_btw_mdl[7,3:5]))` to `r apa_num(max(table_btw_mdl[7,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[8,3:5]))` to `r apa_num(max(table_btw_mdl[8,3:5]))` for unrestrained prediction. Moreover, the three window lengths did not significantly differ for any of the metrics (*p*s \>= `r apa_num(min(table_btw_mdl[1:8,8]))`).

@@

To conclude, the current study provided a powerful tool for quantifying infants’ everyday restraint experiences in the home using wearable sensors and machine learning modeling. The model achieved high accuracy and strong agreement with human annotations, as well as convergent validity by showing an age-related decrease in infant restraint time. We explored the window length in the data aggregating procedure of modeling and proved that a longer window enhanced the model’s balanced performance. Together, the current study advanced the method for measuring infant full-day, real-time movement, and opened avenues for research on how restraint imposed by caregivers shape early development.

# Declarations
## Funding
This work was supported by National Science Foundation Grant BCS #1941449.

## Conflicts of interest
The authors report no conflicts of interest related to this article.

## Ethics approval
All study procedures were reviewed and approved by the Institutional Review Board at the University of California, Riverside (Protocol HS-15-050).

## Consent to participate
Written informed consent was obtained from all caregivers before the study began.

## Consent for publication
Additional written consent was obtained from caregivers to allow the sharing of audio and video data.

## Availability of data and materials
Human annotation results, which were used for model training and validation are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

## Code availability
All the processing scripts are also publicly shared on OSF.

>>>>

abstract: "Physical restraint—including being held, carried, and restrained in devices—is a common feature of infants’ everyday lives. However, previous survey-based and video-based methods cannot simultaneously provide continuous, full-day accounts of infant restraint experiences. This study developed and validated a machine learning model to quantify infants’ restraint time moment-to-moment across the full day in the home environment using wearable sensor data. We used a data set that includes 142 home-visit sessions from 65 infants, with 29 younger infants aged 4-7 months, and 36 older infants aged 11-14 months. We annotated infants’ restrained states in the first 1.5-hour video recordings of each session as ground truth labels. The supervised machine-learning model achieved high accuracy (89%) and substantial kappa agreement (κ = .73) compared with human-coded ground truth. The model presented a slight bias towards overestimating unrestrained periods compared to restrained periods, but the bias was mitigated when we used a longer data-aggregating window length. The model also showed convergent validity by corroborating prior studies that showed an age-related decrease in infants' overall restraint time throughout the day. In short, the current study demonstrated the feasibility of using wearable sensors to quantify infants’ real-world restraint experiences, offering a new tool for studying how daily restraint influences early development."

@@

bibliography: [references.bib, packages.bib]

@@

Physical restraint is a common feature in infants’ everyday lives. Infants are routinely held or carried by caregivers, and infants are often restrained in devices such as highchairs, walkers, and strollers [@Birken2015s]. As an externally regulated aspect of infants’ daily life, restraint shapes infants’ experiences based on caregivers’ decisions--caregivers, not infants, choose when infants are restrained. Restraint limits infants’ physical activities and might hinder infants’ motor development [e.g., @Pin2007z]. Restraint also alters experiences that are relevant for perceptual and cognitive development by determining what is accessible in the physical world. For example, restraint is associated with fewer object holding experiences [@Franchak2024l], reduced amounts of language input [@Malachowski2023a], and increased opportunities for joint attention [@Karasik2025h]. Therefore, characterizing infants’ restrained experiences helps profile infants’ daily learning contexts and situate their motor, perceptual, and cognitive development within everyday experiences.

@@

Despite different approaches, prior work converges to show that restraint is common in infant daily life, and that the time infants spend restrained changes over development. In terms of caregiver restraint, 2- to 6-month-old infants spent around 40% of their awake time being held or carried [@Malachowski2023a; @Siddicky2020f]. Older 10- to 13-month-olds spent around 19% of their awake time being held or carried [@Franchak2024l]. In terms of device restraint, 4- to 6-month-olds spent around 20% of their awake time in seating devices, and 35% in placement with physical support [@Callahan1997s; @Malachowski2023a]. Older 10- to 13-month-olds spent around 27% of awake time in restraining furniture or devices [@Franchak2024l]. Additionally, a study that compared across ages also confirmed that infants’ overall restrained time decreased from 2 to 6 months of age [@Carson2022n].

@@

Researchers have used survey methods, including traditional retrospective report and ecological momentary assessment (EMA), to quantify infants’ restrained time, but survey methods have limitations. Retrospective parent report asks parents to estimate their infant’s average daily restrained time over the past few weeks or month [@Abbott2001u; @Bartlett2003w; @Callahan1997s; @Carson2022n; @Hesketh2015i; @Karasik2018t; @Siddicky2020f]. Despite the widespread use of retrospective reports, they lack precision by relying on parents’ memory—it is difficult to recall and sum every bout of restraint practices and then report an overall time estimation [@Bradburn1987g]. There are exceptions where caregivers were asked to complete structured daily diaries reflecting on the previous day’s restraint practices in short intervals (e.g., every five minutes) [@Karasik2018t; @Majnemer2005j]. However, this method might still be sensitive to memory loss and thus not as accurate in providing data. EMA survey techniques reduce caregivers’ memory burden by sending momentary phone-survey prompts to caregivers multiple times (e.g., 5, 10, or 12 times) randomly throughout a day [@Franchak2019o; @Franchak2024l; @Malachowski2023a]. Despite the benefits of gathering immediate observations rather than retrospective reports, EMA cannot provide continuous time-series data. Thus, a method is needed that provides time-sequence information without relying on parent report.

Video recording and human annotation overcome the above drawbacks of survey-based methods, but video is subject to other limitations. One disadvantage of video observation is being obtrusive. The presence of a researcher in the room may affect the caregiver’s and infant’s behaviors. Another limitation is that video recording and annotation are both labor-intensive and time-consuming, which limits its use in large-scale or long-duration studies. Indeed, prior work that videotaped and manually annotated infants’ restraint had short durations of recording (e.g., half an hour in @Springfield2025p, and one hour in @Wang2025v). Within the short duration of recording, researchers tend to record the period when infants are awake and active. In contrast, periods when the infants are restrained, like seated in front of the TV or seated in highchairs, would likely be skipped. Thus, video recording is subject to sampling bias and might not be representative of infants’ whole-day experiences.

@@

However, one of the challenges researchers continue to face is the reduced model performance in naturalistic home settings [@Airaksinen2024z; @Franchak2024e; @Yao2019q], compared with more controlled laboratory environments [@Airaksinen2020j; @Franchak2021k]. Indeed, infants may exhibit more complex and unexpected movements at home versus in lab, which requires the models to be able to generalize to scenarios that go beyond the models’ training data set. Moreover, since no prior work has used wearable sensors to detect infants’ overall restrained periods, the current study addresses a unique detection problem.

@@

The third step was to train and validate supervised machine learning models that classify the motion features into restrained status. We applied the random forest algorithm (Breiman, 2001) in modeling using the DecisionTree package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as a testing set and trained the model using the other sessions, to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked, and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

@@

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). This reproducible manuscript, including the model evaluation, figures, and statistical analyses, was made with Quarto version 1.5.57 (Posit, PBC, 2024) using R version 4.5.0 [@R-base] with the R packages tidyverse [@R-tidyverse], flextable [@R-flextable], and papaja (Aust & Barth, 2024), as well as the apa quarto template (Schneider, n.d.). Human annotation results, which were used for model training and validation, together with processing scripts are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

@@

```{r fig-cibtwctg}
#| fig-cap: Differences in sensitivity and PPV between restrained and unrestrained prediction according to window length
#| fig-width: 8
#| fig-height: 5
#| apa-note: Each vertical line indicates the 95% CI of the metric's difference between restrained prediction and unrestrained prediction. The dot on each line denotes the mean of the difference (restrained - unrestrained).

@@

@tbl-indexbetweenmodels shows that all window lengths reached high accuracy (`r apa_num(table_btw_mdl[1,3])` for 4 s, `r apa_num(table_btw_mdl[1,4])` for 16 s, and `r apa_num(table_btw_mdl[1,5])` for 30 s), and good kappa values (`r apa_num(table_btw_mdl[2,3])` for 4 s, `r apa_num(table_btw_mdl[2,4])` for 16 s, and `r apa_num(table_btw_mdl[2,5])` for 30 s). All window lengths presented high sensitivity—how often the model correctly detected when the infant was actually restrained or unrestrained—ranging from `r apa_num(min(table_btw_mdl[3,3:5]))` to `r apa_num(max(table_btw_mdl[3,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[4,3:5]))` to `r apa_num(max(table_btw_mdl[4,3:5]))` for unrestrained prediction. All window lengths had high positive predictive value (PPV)—how often the model’s predictions of being restrained or unrestrained were correct—ranging from `r apa_num(min(table_btw_mdl[5,3:5]))` to `r apa_num(max(table_btw_mdl[5,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[6,3:5]))` to `r apa_num(max(table_btw_mdl[6,3:5]))` for unrestrained prediction. All window lengths also had high F1 scores—a balanced measure that combines sensitivity and PPV—ranging from `r apa_num(min(table_btw_mdl[7,3:5]))` to `r apa_num(max(table_btw_mdl[7,3:5]))` for restrained prediction, and `r apa_num(min(table_btw_mdl[8,3:5]))` to `r apa_num(max(table_btw_mdl[8,3:5]))` for unrestrained prediction. Moreover, the three window lengths did not significantly differ for any of the metrics (*p*s $\ge$ `r apa_num(min(table_btw_mdl[1:8,8]))`).

@@

To conclude, the current study provided a powerful tool for quantifying infants’ everyday restraint experiences in the home using wearable sensors and machine learning modeling. The model achieved high accuracy and strong agreement with human annotations, as well as convergent validity by showing an age-related decrease in infant restraint time. We explored the window length in the data aggregating procedure of modeling and proved that a longer window enhanced the model’s balanced performance. Together, the current study advanced the method for measuring infant full-day, real-time movement, and opened avenues for research on how restraint imposed by caregivers can shape early development.

# Declarations

## Funding

This work was supported by National Science Foundation Grant BCS #1941449.

## Conflicts of Interest

The authors report no conflicts of interest related to this article.

## Ethics Approval

All study procedures were reviewed and approved by the Institutional Review Board at the University of California, Riverside (Protocol HS-15-050).

## Consent to Participate

Written informed consent was obtained from all caregivers before the study began.

## Consent for Publication

Additional written consent was obtained from caregivers to allow the sharing of audio and video data.

## Availability of Data and Materials

Human annotation results, which were used for model training and validation are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

## Code Availability

All the processing scripts are also publicly shared on OSF.
